---
title: "Reproducible Research Course Project 1"
author: "Glenn Groves"
date: "22/04/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Download and read the data

```{r}
library(dplyr)
library(ggplot2)
library(lattice)

fileURL = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
dest_file = ".\\data\\repdata_data_activity.zip"
download.file(fileURL,dest_file,method="curl")
Sys.time()

data = read.csv(unz(dest_file,"activity.csv"),header=TRUE,stringsAsFactor=FALSE)
```

# What is mean total number of steps taken per day?  
1. Calculate the total number of steps taken per day  
```{r}
data$date = as.Date(data$date,"%Y-%m-%d")
day = group_by(data,date)
day_total = summarise(day,steps = sum(steps))
```
2. Make a histogram of the total number of steps taken each day  
```{r}
hist(day_total$steps, breaks = 20, main="Frequency of Steps per Day", xlab="Steps per Day")
```

3. Calculate and report the mean and median of the total number of steps taken per day  
```{r}
day_total_mean = round(mean(day_total$steps,na.rm=TRUE),digits=0)
day_total_median = median(day_total$steps,na.rm=TRUE)
day_total_sum = sum(day_total$steps,na.rm=TRUE)

```
*The mean number of total steps per day is `r as.integer(day_total_mean)`. Rounding has been applied.*  
*The median number of total steps per day is `r day_total_median`*  
*(This was not actually requested, however the total number of steps across all days is `r day_total_sum`)*  


# What is the average daily activity pattern?  
1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)  
```{r}
interval = group_by(data,interval)
interval_average = summarise(interval,average_steps = mean(steps,na.rm=TRUE))
plot(interval_average, main="Average Steps for 5 Minute Interval across all Days", xlab="Interval", ylab="Average Steps")
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?  
```{r}
interval_maximum_steps = interval_average[which.max(interval_average$average_steps),]$interval
```
*The interval containing the maximum number of steps, on average across all days, is `r interval_maximum_steps`*  

# Imputing missing values  
1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)  
```{r}
number_missing = sum(is.na(data$steps))
```
*The total number of missing values in the dataset is `r number_missing`*  

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.  
3. Create a new dataset that is equal to the original dataset but with the missing data filled in.  
```{r}
data_imputed = merge(data, interval_average, by="interval")
data_imputed$steps = ifelse(is.na(data_imputed$steps), data_imputed$average_steps, data_imputed$steps)
```

4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.  
```{r}
day_imputed = group_by(data_imputed,date)
day_imputed_total = summarise(day_imputed,steps = sum(steps))
hist(day_imputed_total$steps, breaks = 20, main="Frequency of Steps per Day Including Imputed", xlab="Steps per Day Including Imputed")
imputed_mean_steps = round(mean(day_imputed_total$steps,na.rm=TRUE),digits=0)
imputed_median_steps = round(median(day_imputed_total$steps,na.rm=TRUE),digits=0)
imputed_total_steps = round(sum(day_imputed_total$steps,na.rm=TRUE),digits=0)

```
*The mean number of steps per day, after missing values were replaced with imputed values, is `r as.integer(imputed_mean_steps)`. Rounding has been applied.*  
*(Prior to imputing missing values, the mean number of total steps per day was `r as.integer(day_total_mean)`. Rounding had been applied.)*  
*The median number of steps per day, after missing values were replaced with imputed values, is `r as.integer(imputed_median_steps)`. Rounding has been applied (as an unrounded mean was used to impute missing values).*  
*(Prior to imputing missing values, the median number of total steps per day was `r day_total_median`)*  


Do these values differ from the estimates from the first part of the assignment?  
*Most of the values are the same; however, the occurrence of the most commonly occuring numbers of steps have nearly doubled.*  

What is the impact of imputing missing data on the estimates of the total daily number of steps?  
*The mean and the median total daily numbers of steps are practically unchanged. This makes sense as we used the mean to set the missing values.*  
*The TOTAL number of steps has of course increased. The previous total number of steps was `r day_total_sum`, but after imputing the total number of steps is `r as.integer(imputed_total_steps)`)*  


# Are there differences in activity patterns between weekdays and weekends?  
1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.  
```{r}
day_of_week = mutate(data_imputed, day_type = as.factor(ifelse(weekdays(data_imputed$date) %in% c("Saturday","Sunday"), "weekend","weekday")))
```

2. Make a panel plot containing a time series plot (i.e. type="l") of the 5-minute interval (x-axis) and the average number of steps taken,  
averaged across all weekday days or weekend days (y-axis).  
See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.  
```{r}
day_of_week_average = group_by(day_of_week, interval, day_type)
day_of_week_average = summarise(day_of_week_average,steps = mean(steps))
xyplot(day_of_week_average$steps ~ day_of_week_average$interval | day_of_week_average$day_type, layout=c(1,2), type="l", xlab="Interval", ylab="Number of steps")
```

*Overall the pattern is the same between weekdays and weekends; however, there is a peak of activity around 8:30 a.m. on weekdays that does not occur on the weekend, and the weekend has a slightly higher level of activity from around 10 a.m. to 5 p.m. This would be consistent with travel for weekday activities, e.g. work and school, and is also consistent with the peak seen around 8:30 a.m. in the "Average Steps for 5 Minute Interval across all Days" graph; that peak in the graph would be due to the peak seen at this time in weekdays. As there are 5 weekdays to 2 weekend days, the peak appears to be more pronounced in the "Average Steps for 5 Minute Interval across all Days" graph than when compared to the weekday graph.*  
